---
title: "African Data Analysis"
author: "Ke Ding"
date: "5/23/2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
##Work log
July 3rd:
I addressed the comments from the last meeting and start to write down some of the findings from the graphs we plotted.

```{r, include=FALSE}
getwd()
#setwd("/Users/dr.dk/Desktop/R-programming")
library(ggplot2)
library(reshape2)
library(Hmisc)
require(CCA)
```


```{r}
#remove not useful columns in raw data table
data<-read.csv("WEF_Data_Compilation.csv")
no_use<- names(data) %in% c("X","X.1")
data<-data[!no_use]
```

## WEF Access
```{r}
#make the parameter column as charactor so we can search the certain name in the column
data$Parameter.<-as.character(data$Parameter.)
#search function of names or characters in a data table
#data[grep("access",data$Parameter.,ignore.case=TRUE),]
ac_dw<-subset(data,Parameter.=="Total population with access to safe drinking-water")
#melting dataframes for plotting purposes
ac_dw_melt<-melt(ac_dw,id.vars = c("Parameter.","Unit.","Source.","Year."))
#remove missing values
ac_dw_melt$value<-as.numeric(ac_dw_melt$value)
ac_dw_melt<-ac_dw_melt[complete.cases(ac_dw_melt),]
heading<-names(ac_dw)
ggplot(ac_dw_melt,aes(x=variable,y=value))+ geom_point(aes(color = variable)) +  theme(legend.position="top",
              axis.text=element_text(angle=90, size = 6)) + labs(x="Countries",y= "access to safe drinking-water (%)")+ coord_cartesian(xlim= NULL,ylim = NULL, expand= TRUE)+ geom_text(aes(label=variable),hjust=0, vjust=0)
warning= FALSE
```

```{r}
#data[grep("Depth", data$Parameter., ignore.case=TRUE),]
defi_food<- subset(data, Parameter.=="Depth of the food deficit")
defi_food_melt<- melt(defi_food,id.vars= c("Parameter.","Unit.","Source.","Year."))
defi_food_melt$value<- as.numeric(defi_food_melt$value)
defi_food_melt<- defi_food_melt[complete.cases(defi_food_melt),]
heading<-names(defi_food_melt)
ggplot(defi_food_melt,aes(x=variable,y=value))+ geom_point(aes(color = variable)) + coord_cartesian(xlim= NULL,ylim = NULL, expand= TRUE)+  theme(legend.position="top",
              axis.text=element_text(angle=90, size = 6))+ geom_text(aes(label=variable),hjust=0, vjust=0)+ labs(x="Countries",y= "kcal/capita/day, 3yr avg",title= "Depth of the Food Deficit")
```
```{r}
#data[grep("energy supply", data$Parameter., ignore.case=TRUE),]
ener_pp<- subset(data, Parameter.=="Energy supply per capita")
ener_pp_melt<- melt(ener_pp,id.vars= c("Parameter.","Unit.","Source.","Year."))
ener_pp_melt$value<- as.numeric(ener_pp_melt$value)
ener_pp_melt<- ener_pp_melt[complete.cases(ener_pp_melt),]
heading<-names(ener_pp_melt)
ggplot(ener_pp_melt,aes(x=variable,y=value))+ geom_point(aes(color = variable)) + coord_cartesian(xlim= NULL,ylim = NULL, expand= TRUE)+  theme(legend.position="top",
              axis.text=element_text(angle=90, size = 6))+ labs(x="Countries",y= "Gigajoules",title= "Energy Supply Per Capita")+ geom_text(aes(label=variable),hjust=0, vjust=0)
```
```{r}
ac_dw_1<-melt(ac_dw,id.vars= c("Parameter.","Unit.","Source.","Year."))
ac_dw_data<-names(ac_dw_1) %in% c("variable","value")
ac_dw_c<-ac_dw_1[ac_dw_data]
colnames(ac_dw_c)<-c("Countries","Total population with access to safe drinking-water")
defi_food_1<-melt(defi_food,id.vars= c("Parameter.","Unit.","Source.","Year."))
defi_food_data<-names(ac_dw_1) %in% c("value")
defi_food_c<-defi_food_1[defi_food_data]
colnames(defi_food_c)<-c("Depth of the food deficit")

#Merge columns of interest
ac_dw_defi_food<-cbind(ac_dw_c,defi_food_c)
colnames(ac_dw_defi_food)<-c("Countries","ac_safe_dw","defi_food") #rename  columns for ease

#format columns appropriately
ac_dw_defi_food$Countries<- as.character(ac_dw_defi_food$Countries)
ac_dw_defi_food$ac_safe_dw<- as.numeric(ac_dw_defi_food$ac_safe_dw)
ac_dw_defi_food$defi_food<- as.numeric(ac_dw_defi_food$defi_food)

#subsetting 5 reference countries
ac_dw_defi_food_5c<-tail(ac_dw_defi_food,5) 
ac_dw_defi_food_af<-head(ac_dw_defi_food,-5)

#if needed, can do complete cases
#ac_dw_defi_food<- ac_dw_defi_food[-c(1,16),]
#complete.cases(ac_dw_defi_food)

ggplot(ac_dw_defi_food_af,aes(x=ac_safe_dw, y=defi_food, color=Countries))+ 
  geom_point() + geom_smooth(method=lm,color="black")+ 
  labs(x="access to safe drinking water (%)",y= "deficit of food supply (kcal/capita/day, 3yr avg)") +
  theme_bw()+ geom_text(aes(label=Countries),hjust=0, vjust=0)
```
```{r}
#subset data of energy per capita
ener_pp_1<-melt(ener_pp,id.vars= c("Parameter.","Unit.","Source.","Year."))
#subset the value data only
ener_pp_data<-names(ener_pp_1) %in% c("value")
#_c means clean (data only)
ener_pp_c<-ener_pp_1[ener_pp_data]
colnames(defi_food_c)<-c("Depth of the food deficit")

#Merge columns of interest
ac_dw_ener_pp<-cbind(ac_dw_c,ener_pp_c)
colnames(ac_dw_ener_pp)<-c("Countries","ac_safe_dw","ener_pp") #rename  columns for ease


#format columns appropriately
ac_dw_ener_pp$Countries<- as.character(ac_dw_ener_pp$Countries)
ac_dw_ener_pp$ac_safe_dw<- as.numeric(ac_dw_ener_pp$ac_safe_dw)
ac_dw_ener_pp$ener_pp<- as.numeric(ac_dw_ener_pp$ener_pp)

#subsetting 5 reference countries
ac_dw_ener_pp_5c<-tail(ac_dw_ener_pp,5) 
ac_dw_ener_pp_af<-head(ac_dw_ener_pp,-5)

#remove outliers
ac_dw_ener_pp_af<-ac_dw_ener_pp_af[-c(5,6),]

#if needed, can do complete cases
#ac_dw_defi_food<- ac_dw_defi_food[-c(1,16),]
#complete.cases(ac_dw_defi_food)
ggplot(ac_dw_ener_pp_af, aes(x=ac_safe_dw, y=ener_pp, color=Countries))+ 
  geom_point() +stat_smooth(method=lm,color="black")+
  labs(x="access to safe drinking water (%)",y= "Energy supply per capita (Gigajoules)") +
  theme_bw()+ geom_text(aes(label=Countries),hjust=0, vjust=0)
```
```{r}

```

```{r}

#Merge columns of interest
defi_food_ener_pp<-cbind(defi_food_c,ener_pp_c)
defi_food_ener_pp<-cbind(ac_dw_c$Countries,defi_food_ener_pp)
colnames(defi_food_ener_pp)<-c("Countries","defi_food","ener_pp") #rename  columns for ease
#format columns appropriately
defi_food_ener_pp$Countries<- as.character(defi_food_ener_pp$Countries)
defi_food_ener_pp$defi_food<- as.numeric(defi_food_ener_pp$defi_food)
defi_food_ener_pp$ener_pp<- as.numeric(defi_food_ener_pp$ener_pp)

#subsetting 5 reference countries
defi_food_ener_pp_5c<-tail(defi_food_ener_pp,5)
defi_food_ener_pp_af<- head(defi_food_ener_pp,-5)
df_ep_countries<-defi_food_ener_pp_af$Countries


#if needed, can do complete cases
#ac_dw_defi_food<- ac_dw_defi_food[-c(1,16),]
#complete.cases(ac_dw_defi_food)
ggplot(defi_food_ener_pp_af, aes(x=defi_food, y=ener_pp, color=df_ep_countries))+
  geom_point()+ geom_smooth(method="lm", color="black") +
  labs(x="deficit of food supply (kcal/capita/day, 3yr avg)", y= "Energy supply per capita (Gigajoules)")+
  theme_bw()+ geom_text(aes(label=df_ep_countries),hjust=0, vjust=0,size=2)
```



## Correlation analysis
The series of correlation analysis includes: correlations analysis, principle component analysis, and canonical correlation analysis.


```{r}

#form a matrix of all three access parameters
ac_3<- cbind(ac_dw_c[,2],ener_pp_c,defi_food_c)
colnames(ac_3)<- c("ac_dw","ener_pp","defi_food")
row.names(ac_3)<- ac_dw_defi_food$Countries
#format the matrix to characters and numeric
#ac_3$Countries<- as.character(ac_3$Countries)
ac_3$ac_dw<- as.numeric(ac_3$ac_dw)
ac_3$defi_food<- as.numeric(ac_3$defi_food)
ac_3$ener_pp<- as.numeric(ac_3$ener_pp)
ac_3_5countries<-ac_3[c(15,16,17,18,19),]
ac_3_africa<-ac_3[-c(15,16,17,18,19),]
ac_3<- ac_3[complete.cases(ac_3),]
ac_3_africa<- ac_3_africa[complete.cases(ac_3_africa),]
print("Correlations of Access to WEF resources for african countries and 5 reference countries")
rcorr(as.matrix(ac_3))
print('-------------------------')
print("correlations of WEF Resources just for african countries")
rcorr(as.matrix(ac_3_africa))
```

```{r}
#principle component analysis on access
library(HSAUR)
ac_3_pca<- prcomp(ac_3_africa,center=TRUE, scale = TRUE)
print("Principal component analysis on access of WEF")
print(ac_3_pca)
summary(ac_3_pca)
screeplot(ac_3_pca, type="lines")
a1<- ac_3_pca$rotation[,1]
a1
```
#get access scores for each country: Country Para*PC1
```{r}
#get access scores for each country: Country Para*PC
score_pc1_ac_3<-a1*ac_3
print(score_pc1_ac_3)
```

```{r}
#subset health data all selected data are all in %
diarrhea<-subset(data,Parameter.=="Diarrhea as a cause of death for children under 5")
death_airpollu<- subset(data,Parameter.=="Causes of Death attributable to Air Pollt")
perc_stunt<- subset(data,Parameter.=="Percentage of children under 5 years of age who are stunted")
perc_wast<- subset(data,Parameter.=="Percentage of children under 5 years of age affected by wasting")
perc_underw<- subset(data,Parameter.=="Percentage of children under 5 years of age who are underweight")
#extract values
diarrhea_melt<-melt(diarrhea,id.vars= c("Parameter.","Unit.","Source.","Year."))
diarrhea_melt$value<-as.numeric(diarrhea_melt$value)

death_airpollu_melt<-melt(death_airpollu,id.vars= c("Parameter.","Unit.","Source.","Year."))
death_airpollu_melt$value<-as.numeric(death_airpollu_melt$value)

# perc_stunt_melt<-melt(perc_stunt,id.vars= c("Parameter.","Unit.","Source.","Year."))
# perc_stunt_melt$value<-as.numeric(perc_stunt_melt$value)
# 
# perc_wast_melt<-melt(perc_wast,id.vars= c("Parameter.","Unit.","Source.","Year."))
# perc_wast_melt$value<-as.numeric(perc_wast_melt$value)

perc_underw_melt<-melt(perc_underw,id.vars= c("Parameter.","Unit.","Source.","Year."))
perc_underw_melt$value<-as.numeric(perc_underw_melt$value)

# create new data frame of health parameters
#more detailed one
# health<- cbind(diarrhea_melt$value, death_airpollu_melt$value, perc_stunt_melt$value, perc_wast_melt$value, perc_underw_melt$value) 
#simple one
health<- cbind(diarrhea_melt$value, death_airpollu_melt$value,  perc_underw_melt$value) 

#format matrix change column names and row names
health<-as.data.frame(health)

#colnames(health)<-c("diarrhea","death_airpollu","U5stunt","U5wast","U5underw")
#create a string with all names of countries
Countries<-defi_food_ener_pp$Countries
colnames(health)<-c("diarrhea","death_airpollu","U5underw")
row.names(health)<-Countries
health<- health[complete.cases(health),]
```
#correlation analysis on Health
```{r}
#correlation analysis on Health
rcorr(as.matrix(health))
```
#PCA on Health Parameters
```{r}
#PCA on health 
health_pca<- prcomp(health,center=TRUE, scale = TRUE)
print("PCA on Health parameters")
print(health_pca)
summary(health_pca)
screeplot(health_pca, type="lines")
health_1<- health_pca$rotation[,1]
health_1
```
#scores of health parameters: health*pc1
```{r}
#scores of health parameters: health*pc1
score_pc1_health<-health_1*health
print(score_pc1_health)
```

```{r}
ggplot(health, aes(x=diarrhea, y=U5underw, color=row.names(health)))+
  geom_point()+ geom_smooth(method="lm", color="black") +
  labs(x="diarrhea (%)", y= "U5underweight (%)",title= "Percentage of children under 5 die due to diarrhea vs children under 5 who are underweight ")+
  theme_bw()+geom_text(label=row.names(health),hjust=0,vjust=0,size=2)
```

```{r}
#subset equality parameters
G_inequal<-subset(data,Parameter.=="Gender inequality index")
G_inequal_melt<-melt(G_inequal,id.vars= c("Parameter.","Unit.","Source.","Year."))
G_inequal_melt$value<-as.numeric(G_inequal_melt$value)

#rural parameters
#safe drinkingwater
ac_dw_ru<-subset(data,Parameter.=="Rural population with access to safe drinking-water")
ac_dw_ru_melt<-melt(ac_dw_ru,id.vars=c("Parameter.","Unit.","Source.","Year."))
ac_dw_ru_melt$value<-as.numeric(ac_dw_ru_melt$value)
#clean cooking
ac_cc_ru<-subset(data,Parameter.=="Access to clean cooking in rural areas")
ac_cc_ru_melt<-melt(ac_cc_ru,id.vars=c("Parameter.","Unit.","Source.","Year."))
ac_cc_ru_melt$value<-as.numeric(ac_cc_ru_melt$value)

#urban parameters
#safe drinkingwater
ac_dw_ur<-subset(data,Parameter.=="Urban population with access to safe drinking-water")
ac_dw_ur_melt<-melt(ac_dw_ur,id.vars=c("Parameter.","Unit.","Source.","Year."))
ac_dw_ur_melt$value<-as.numeric(ac_dw_ur_melt$value)
#clean cooking
ac_cc_ur<-subset(data,Parameter.=="Access to clean cooking in urban areas")
ac_cc_ur_melt<-melt(ac_cc_ur,id.vars=c("Parameter.","Unit.","Source.","Year."))
ac_cc_ur_melt$value<-as.numeric(ac_cc_ur_melt$value)

#combine as a equality subset
equality<-cbind(G_inequal_melt$value, ac_dw_ru_melt$value, ac_cc_ru_melt$value, ac_dw_ur_melt$value, ac_cc_ur_melt$value)
equality<-as.data.frame(equality)
colnames(equality)<-c("G_inequal","ac_dw_rural","ac_cc_rural","ac_dw_urban","ac_cc_urban")
row.names(equality)<-Countries
equality_af<-head(equality,-5)
```

```{r}
ggplot(equality_af, aes(x=ac_dw_urban, y=ac_dw_rural, color=row.names(equality_af)))+
  geom_point()+ geom_smooth(method="lm", color="black") +
  labs(x="Urban population with access to safe drinking-water (%)", y= "rural population with access to safe drinking-water (%)", title= "Percentage of Rural and Urban population who have access to safe drinking water")+
  theme_bw()+geom_text(label=row.names(equality_af),hjust=0,vjust=0,size=2)
```

```{r}
ggplot(equality, aes(x=ac_cc_urban, y=ac_cc_rural, color=Countries))+
  geom_point()+ geom_smooth(method="lm", color="black") +
  labs(x="Urban population with access to clean cooking (%)", y= "rural population with access to clean cooking (%)", title= "Percentage of Rural and Urban population who have access to clean cooking")+
  theme_bw()+geom_text(label=Countries,hjust=0,vjust=0,size=2)
```

```{r}
#subsetting data of urban and rural population
ur_pop<-subset(data,Parameter.=="Urban Population")
ru_pop<-subset(data,Parameter.=="Rural Population")

#melting dataframes for plotting purposes
ur_pop_melt<-melt(ur_pop,id.vars = c("Parameter.","Unit.","Source.","Year."))
ur_pop_melt$value<-as.numeric(ur_pop_melt$value)
ru_pop_melt<-melt(ru_pop,id.vars = c("Parameter.","Unit.","Source.","Year."))
ru_pop_melt$value<-as.numeric(ru_pop_melt$value)

ur_pop_c<-ur_pop_melt$value
ru_pop_c<-ru_pop_melt$value
pop_countries<-ur_pop_melt$variable
ur_ru_pop<-cbind(ur_pop_c,ru_pop_c)
colnames(ur_ru_pop)<-c("urban_pop","rural_pop")
row.names(ur_ru_pop)<-pop_countries
ur_ru_pop<-as.data.frame(ur_ru_pop)
ur_ru_pop_af<-head(ur_ru_pop,-5)
ur_ru_pop_5ref<-tail(ur_ru_pop,5)
```

```{r}
ggplot(ur_ru_pop_af, aes(x=urban_pop, y=rural_pop, color=row.names(ur_ru_pop_af)))+
  geom_point()+ geom_smooth(method="lm", color="black") +
  labs(x="Urban population (1000 inhabitants)", y= "Rural population (1000 habitants)", title="Urban population vs Rural Population in African Countries")+
  theme_bw()+geom_text(label=row.names(ur_ru_pop_af),hjust=0,vjust=0,size=2)
```
```{r}
ggplot(ur_ru_pop_5ref, aes(x=urban_pop, y=rural_pop, color=row.names(ur_ru_pop_5ref)))+
  geom_point()+ geom_smooth(method="lm", color="black") +
  labs(x="Urban population (1000 inhabitants)", y= "Rural population (1000 habitants)",title="Urban population vs Rural Population in 5 reference countries")+
  theme_bw()+geom_text(label=row.names(ur_ru_pop_5ref),hjust=0,vjust=0,size=2)
```

#correlation in equality
```{r}
#correlation in equality
#equality<- equality[complete.cases(equality),]
print("Correlations in Equality parameters")
rcorr(as.matrix(equality))
```
#principle component analysis in equality
```{r}
#principle component analysis in equality
equality<- equality[complete.cases(equality),]
equality_pca<- prcomp(equality,center=TRUE, scale = TRUE)

print("PCA on Equality")
print(equality_pca)
summary(equality_pca)
screeplot(equality_pca, type="lines")
equality_1<- equality_pca$rotation[,1]
equality_1
```
#correlation on the first PC of health and access

```{r}
#correlation on the first PC of health and access
#clean two data sets to perform correlation analysis

score_pc1_health_c<-score_pc1_health[!rownames(score_pc1_health)%in% "Equatorial.Guinea",]
matcor(score_pc1_health_c, score_pc1_ac_3)
```
#Canonical Correlation Analysis
```{r}

print("Correlations between access set and health set")
matcor(ac_3[1:12,], health[1:12,])
print("Canonical Correlation Analysis on Access and Health")
cc1<-cc(ac_3[1:12,], health[1:12,])
print("_____________________________")
print(cc1)
```
