---
title: "SSA data Analysis"
author: "Ke Ding"
date: "8/23/2017"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

##Work log
July 3rd:
I addressed the comments from the last meeting and start to write down some of the findings from the graphs we plotted.

July 19th:
I expanded my database and start to do the data analysis.

August 18th:
I filled the data in the access matrix and independent matrix and performed the pca analyses for the two matrix.

August 27th:
I added the bar plots for the each variable in the independent matrix.
I also added the plot of HDI vs. Seasonal Variability.

```{r, include=FALSE}
getwd()
#setwd("/Users/dr.dk/Desktop/R-programming")
library(ggplot2)
library(reshape2)
library(Hmisc)
library(ggrepel)
require(CCA)
library(cluster)
library(fpc)
options(warn=-1)
```


```{r}
#remove not useful columns in raw data table
data<-read.csv("WEF_Data_Compilation_ALL_SSA.csv")
no_use<- names(data) %in% c("X")
data<-data[!no_use]
```

## WEF Access

#Total Population with access to safe drinking water

The range of percentage of total population with access to safe drinking water varies from 50% to nearly 100%. African Countires such as Mauritius, Sao Tome and Principle, and Botswana performed very well, however, 8 out of 49 SSA countries were around the bottom line of 
50%. The average percentage of access to safe drinking water is close to 80% for all SSA countries.

```{r}
#make the parameter column as charactor so we can search the certain name in the column
data$Parameter.<-as.character(data$Parameter.)
#search function of names or characters in a data table
#data[grep("access",data$Parameter.,ignore.case=TRUE),]
ac_dw<-subset(data,Parameter.=="Total population with access to safe drinking-water")
#melting dataframes for plotting purposes
ac_dw_melt<-melt(ac_dw,id.vars = c("Parameter.","Unit.","Source.","Year."))
#remove missing values
ac_dw_melt$value<-as.numeric(ac_dw_melt$value)
ac_dw_melt<-ac_dw_melt[complete.cases(ac_dw_melt),]
heading<-names(ac_dw)
ac_dw_melt_ssa<-head(ac_dw_melt[,-5])
avg_ac_dw<-mean(ac_dw_melt_ssa$value)
ggplot(ac_dw_melt,aes(x=variable,y=value))+ geom_col(mapping = NULL, data = NULL, position = "stack",
  width = NULL, na.rm = FALSE, show.legend = NA, inherit.aes = TRUE)+ theme(legend.position="top",
              axis.text=element_text(angle=90, size = 6))+ labs(x="Countries", y= "access to safe drinking-water (%)")+ coord_cartesian(xlim= NULL,ylim = NULL, expand= TRUE)+geom_text(aes(label=variable),size=2,hjust=0.5,vjust=0)+geom_hline(yintercept = avg_ac_dw)
warning= FALSE

```

#Depth of The Food Deficit

The depth of food deficit, measured by calories per person per day, varies from 20 calories to 400 calories, with an average of 140kcal/person/day. Countries such as Central African Republic, Namibia and Zambia exceeded 300 calories of food deficit daily. In the other hand, countries such as Ghana, Gabon, Gambia and South Africa nearly have no deficit in food. 

```{r}
#data[grep("Depth", data$Parameter., ignore.case=TRUE),]
defi_food<- subset(data, Parameter.=="Depth of the food deficit")
defi_food_melt<- melt(defi_food,id.vars= c("Parameter.","Unit.","Source.","Year."))
defi_food_melt$value<- as.numeric(defi_food_melt$value)
defi_food_melt<- defi_food_melt[complete.cases(defi_food_melt),]
heading<-names(defi_food_melt)
avg_defi_food<-mean(defi_food_melt$value)
ggplot(defi_food_melt,aes(x=variable,y=value))+geom_col(mapping = NULL, data = NULL, position = "stack",
  width = NULL, na.rm = FALSE, show.legend = NA, inherit.aes = TRUE) + coord_cartesian(xlim= NULL,ylim = NULL, expand= TRUE)+  theme(legend.position="top",
              axis.text=element_text(angle=90, size = 6))+ labs(x="Countries",y= "kcal/capita/day, 3yr avg",title= "Depth of the Food Deficit")+geom_text(aes(label=variable),size=2,hjust=0.5,vjust=0)+geom_hline(yintercept = avg_defi_food)
warning= FALSE
```

# Energy Supply

Energy supplies measured in Gigajoules per capita in SSA countires indicated that majority of the SSA countries still utilize very little energy, which may imply the relatively low utilization of electricity and living condition. Countries such as Equatorial Guinea, Seychelles, Gabon, and South Africa are outliers which utilized more than double of the average energy supplies of the SSA countries. There are 15 countries ultilize less than 10 GJ per person. 

```{r}
#data[grep("energy supply", data$Parameter., ignore.case=TRUE),]
ener_pp<- subset(data, Parameter.=="Energy supply per capita")
ener_pp_melt<- melt(ener_pp,id.vars= c("Parameter.","Unit.","Source.","Year."))
ener_pp_melt$value<- as.numeric(ener_pp_melt$value)
ener_pp_melt<- ener_pp_melt[complete.cases(ener_pp_melt),]
heading<-names(ener_pp_melt)
avg_energ_pp<-mean(ener_pp_melt$value)
ggplot(ener_pp_melt,aes(x=variable,y=value))+geom_col(mapping = NULL, data = NULL, position = "stack",
  width = NULL, na.rm = FALSE, show.legend = NA, inherit.aes = TRUE) + coord_cartesian(xlim= NULL,ylim = NULL, expand= TRUE)+  theme(legend.position="top",
              axis.text=element_text(angle=90, size = 6))+ labs(x="Countries",y= "Gigajoules",title= "Energy Supply Per Capita")+geom_text(aes(label=variable),size=2,hjust=0.5,vjust=0)+ geom_hline(yintercept = avg_energ_pp)

warning= FALSE
```

# Access to Safe Drinkingwater vs. Food Deficit

```{r}
ac_dw_1<-melt(ac_dw,id.vars= c("Parameter.","Unit.","Source.","Year."))
ac_dw_data<-names(ac_dw_1) %in% c("variable","value")
ac_dw_c<-ac_dw_1[ac_dw_data]
colnames(ac_dw_c)<-c("Countries","Total population with access to safe drinking-water")
defi_food_1<-melt(defi_food,id.vars= c("Parameter.","Unit.","Source.","Year."))
defi_food_data<-names(ac_dw_1) %in% c("value")
defi_food_c<-defi_food_1[defi_food_data]
colnames(defi_food_c)<-c("Depth of the food deficit")

#Merge columns of interest
ac_dw_defi_food<-cbind(ac_dw_c,defi_food_c)
colnames(ac_dw_defi_food)<-c("Countries","ac_safe_dw","defi_food") #rename  columns for ease

#format columns appropriately
ac_dw_defi_food$Countries<- as.character(ac_dw_defi_food$Countries)
ac_dw_defi_food$ac_safe_dw<- as.numeric(ac_dw_defi_food$ac_safe_dw)
ac_dw_defi_food$defi_food<- as.numeric(ac_dw_defi_food$defi_food)

#subsetting 5 reference countries
ac_dw_defi_food_5c<-tail(ac_dw_defi_food,5) 
ac_dw_defi_food_af<-head(ac_dw_defi_food,-5)

#if needed, can do complete cases
#ac_dw_defi_food<- ac_dw_defi_food[-c(1,16),]
#complete.cases(ac_dw_defi_food)

ggplot(ac_dw_defi_food_af,aes(x=ac_safe_dw, y=defi_food, color=Countries))+ 
  geom_point() + geom_smooth(method=lm,color="black")+ 
  labs(x="access to safe drinking water (%)",y= "deficit of food supply (kcal/capita/day, 3yr avg)") +theme(legend.position = "none")+ geom_text_repel(aes(ac_safe_dw, defi_food, label = Countries))
#Try ggrepel for better graphical presentation
# ggplot(ac_dw_defi_food_af) +
#   geom_point(aes(ac_safe_dw, defi_food), color = 'Countries') +
#   geom_text_repel(aes(wt, mpg, label = rownames(dat))) +
#   theme_classic(base_size = 16)
warning= FALSE
```

Countries such as Zambia, Central Africa Republic, Namibia are the most significant outliers which face serious issues of food deficit. 
Angola, Mauritania, and Niger are outliers which have poor accessibility to safe drinkingwater but relatively moderate in food deficit.
Note that, Mauritania and Niger are partially within the Sahara. 

#Energy supply vs. Access to Safe Drinkingwater

```{r}
#subset data of energy per capita
ener_pp_1<-melt(ener_pp,id.vars= c("Parameter.","Unit.","Source.","Year."))
#subset the value data only
ener_pp_data<-names(ener_pp_1) %in% c("value")
#_c means clean (data only)
ener_pp_c<-ener_pp_1[ener_pp_data]
colnames(defi_food_c)<-c("Depth of the food deficit")

#Merge columns of interest
ac_dw_ener_pp<-cbind(ac_dw_c,ener_pp_c)
colnames(ac_dw_ener_pp)<-c("Countries","ac_safe_dw","ener_pp") #rename  columns for ease


#format columns appropriately
ac_dw_ener_pp$Countries<- as.character(ac_dw_ener_pp$Countries)
ac_dw_ener_pp$ac_safe_dw<- as.numeric(ac_dw_ener_pp$ac_safe_dw)
ac_dw_ener_pp$ener_pp<- as.numeric(ac_dw_ener_pp$ener_pp)

#subsetting 5 reference countries
ac_dw_ener_pp_5c<-tail(ac_dw_ener_pp,5) 
ac_dw_ener_pp_af<-head(ac_dw_ener_pp,-5)

#remove outliers
ac_dw_ener_pp_af<-ac_dw_ener_pp_af[-c(5,6),]

#if needed, can do complete cases
#ac_dw_defi_food<- ac_dw_defi_food[-c(1,16),]
#complete.cases(ac_dw_defi_food)
ggplot(ac_dw_ener_pp_af, aes(x=ac_safe_dw, y=ener_pp, color=Countries))+ 
  geom_point() +stat_smooth(method=lm,color="black")+
  labs(x="access to safe drinking water (%)",y= "Energy supply per capita (Gigajoules)")+theme(legend.position = "none")+ geom_text_repel(aes(ac_safe_dw, ener_pp, label = Countries))
warning= FALSE
```

The energy supply versus access to safe drinkingwater plot reveals that despite the outliers such as South Africa, Equatorial Guinea, Seychelles, the variances are more in access to safe drinkingwater rather than energy supplies for SSA countries.

#Food deficit vs. Energy supply

```{r}

#Merge columns of interest
defi_food_ener_pp<-cbind(defi_food_c,ener_pp_c)
defi_food_ener_pp<-cbind(ac_dw_c$Countries,defi_food_ener_pp)
colnames(defi_food_ener_pp)<-c("Countries","defi_food","ener_pp") #rename  columns for ease
#format columns appropriately
defi_food_ener_pp$Countries<- as.character(defi_food_ener_pp$Countries)
defi_food_ener_pp$defi_food<- as.numeric(defi_food_ener_pp$defi_food)
defi_food_ener_pp$ener_pp<- as.numeric(defi_food_ener_pp$ener_pp)

#subsetting 5 reference countries
defi_food_ener_pp_5c<-tail(defi_food_ener_pp,5)
defi_food_ener_pp_af<- head(defi_food_ener_pp,-5)
df_ep_countries<-defi_food_ener_pp_af$Countries

# x<-defi_food_ener_pp_af$defi_food
# y<-defi_food_ener_pp_af$ener_pp
# lm_eqn <- function(defi_food_ener_pp_af){
#     m <- lm(y ~ x, defi_food_ener_pp_af);
#     eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2, 
#          list(a = format(coef(m)[1], digits = 2), 
#               b = format(coef(m)[2], digits = 2), 
#              r2 = format(summary(m)$r.squared, digits = 3)))
#     as.character(as.expression(eq));                 
# }
# print(lm_eqn(defi_food_ener_pp_af))
#if needed, can do complete cases
#ac_dw_defi_food<- ac_dw_defi_food[-c(1,16),]
#complete.cases(ac_dw_defi_food)
ggplot(defi_food_ener_pp_af, aes(x=defi_food, y=ener_pp, color=df_ep_countries))+
  geom_point()+ geom_smooth(method="lm", color="black") +
  labs(x="deficit of food supply (kcal/capita/day, 3yr avg)", y= "Energy supply per capita (Gigajoules)")+
  theme(legend.position = "none")+ geom_text_repel(aes(defi_food, ener_pp, label = Countries))
warning= FALSE
```

Most obviously, South Africa is the outlier which is not suprising. Mozambique and Lesotho are the two countries wrapped by South Africa, relatively closed to each other but varied significantly in deficit of food supply.

##Correlation PCA and CCA statistical Analysis

```{r}
#form a matrix of all three access parameters
ac_3<- cbind(ac_dw_c[,2],ener_pp_c,defi_food_c)
colnames(ac_3)<- c("ac_dw","ener_pp","defi_food")
row.names(ac_3)<- ac_dw_defi_food$Countries
#format the matrix to characters and numeric
#ac_3$Countries<- as.character(ac_3$Countries)
ac_3$ac_dw<- as.numeric(ac_3$ac_dw)
ac_3$defi_food<- as.numeric(ac_3$defi_food)
ac_3$ener_pp<- as.numeric(ac_3$ener_pp)
ac_3_5countries<-tail(ac_3,5)
ac_3_africa<-head(ac_3,-5)
ac_3<- ac_3[complete.cases(ac_3),]
ac_3_africa<- ac_3_africa[complete.cases(ac_3_africa),]
# print("Correlations of Access to WEF resources for african countries and 5 reference countries")
# rcorr(as.matrix(ac_3))
# print('-------------------------')
print("correlations of WEF Resources for SSA countries")
rcorr(as.matrix(ac_3_africa))
warning= FALSE
```

When we added more countries in our analysis, the correlation coefficient dropped. The ac_dw vs. ener_pp dropped from 0.57 to 0.34, and ac_dw vs. defi_food dropped from -0.70 to -0.32. P values suggested that there are significant statistitical correlations (p=0.0376) between ac_dw and ener_pp, while ac_dw and defi_food have less strong statistical correlations (p=0.0518).

#PCA on WEF Access

```{r}
#principle component analysis on access
library(HSAUR)
ac_3_pca<- prcomp(ac_3_africa,center=TRUE, scale = TRUE)
print("Principal component analysis on access of WEF")
print(ac_3_pca)
summary(ac_3_pca)
screeplot(ac_3_pca, type="lines")
a1<- ac_3_pca$rotation[,1]
print(a1)
warning= FALSE
```

From the principal component analysis, the first two pc accounts for nearly 80% of the variances of the matrix. 

```{r}
#get access scores for each country: Country Para*PC
score_pc1_ac_3<-a1*ac_3
#print(score_pc1_ac_3)
warning= FALSE
```

```{r access pc1 plot}
sum_score_pc1_ac_3<-rowSums(score_pc1_ac_3[,1:3])
score_pc1_ac_3["Sum_pc1"]<-sum_score_pc1_ac_3
ggplot(score_pc1_ac_3,aes(x=row.names(score_pc1_ac_3),y=Sum_pc1), warning= FALSE)+ geom_col(mapping = NULL, data = NULL, position = "stack",
  width = NULL, na.rm = FALSE, show.legend = NA, inherit.aes = TRUE)+theme_bw()+ theme(legend.position="top",
              axis.text=element_text(angle=90, size = 6))+ labs(x="Countries", y= "scores of PC1")+ coord_cartesian(xlim= NULL,ylim = NULL, expand= TRUE)+geom_text(aes(label=row.names(score_pc1_ac_3)),size=2,hjust=0.5,vjust=0)


```
This plot is the scores of first pc of WEF Accessibility Matrix for each SSA country. SUM(pc1 * each variable)


```{r}
#subset health data all selected data are all in %
diarrhea<-subset(data,Parameter.=="Diarrhea as a cause of death for children under 5")
death_airpollu<- subset(data,Parameter.=="Causes of Death attributable to Air Pollt")
perc_stunt<- subset(data,Parameter.=="Percentage of children under 5 years of age who are stunted")
perc_wast<- subset(data,Parameter.=="Percentage of children under 5 years of age affected by wasting")
perc_underw<- subset(data,Parameter.=="Percentage of children under 5 years of age who are underweight")
#extract values
diarrhea_melt<-melt(diarrhea,id.vars= c("Parameter.","Unit.","Source.","Year."))
diarrhea_melt$value<-as.numeric(diarrhea_melt$value)

death_airpollu_melt<-melt(death_airpollu,id.vars= c("Parameter.","Unit.","Source.","Year."))
death_airpollu_melt$value<-as.numeric(death_airpollu_melt$value)

# perc_stunt_melt<-melt(perc_stunt,id.vars= c("Parameter.","Unit.","Source.","Year."))
# perc_stunt_melt$value<-as.numeric(perc_stunt_melt$value)
# 
# perc_wast_melt<-melt(perc_wast,id.vars= c("Parameter.","Unit.","Source.","Year."))
# perc_wast_melt$value<-as.numeric(perc_wast_melt$value)

perc_underw_melt<-melt(perc_underw,id.vars= c("Parameter.","Unit.","Source.","Year."))
perc_underw_melt$value<-as.numeric(perc_underw_melt$value)

# create new data frame of health parameters
#more detailed one
# health<- cbind(diarrhea_melt$value, death_airpollu_melt$value, perc_stunt_melt$value, perc_wast_melt$value, perc_underw_melt$value) 
#simple one
health<- cbind(diarrhea_melt$value, death_airpollu_melt$value,  perc_underw_melt$value) 

#format matrix change column names and row names
health<-as.data.frame(health)

#colnames(health)<-c("diarrhea","death_airpollu","U5stunt","U5wast","U5underw")
#create a string with all names of countries
Countries<-defi_food_ener_pp$Countries
colnames(health)<-c("diarrhea","death_airpollu","U5underw")
row.names(health)<-Countries
health<- health[complete.cases(health),]
warning= FALSE
```
#Correlation analysis on Health Parameters

```{r}
#correlation analysis on Health
rcorr(as.matrix(health))
warning= FALSE
```
Unlike the analysis for 14 SSA countries, Diarrhea and underweight issues for Children under 5 have significant statistical correlation
but the diarrhea and air pollution shows much weaker correlation. 
We discussed to drop other two U5 parameters for convenience, however, such change by data expansion suggests we might need to include the other two U5 parameters as well.
```{r}
countries_health<-row.names(health)
countries_health<-as.character(countries_health)
ggplot(health, aes(x=health$diarrhea, y=health$U5underw, color=countries_health))+
  geom_point()+ geom_smooth(method="lm", color="black") +
  labs(x="Percentage of Diarrhea", y= "Percentage of Underweight for Children U5")+
  theme(legend.position = "none")+ geom_text_repel(aes(diarrhea, U5underw, label = countries_health))
warning= FALSE
```

#PCA on Health Parameters

```{r}
#PCA on health 
health_pca<- prcomp(health,center=TRUE, scale = TRUE)
print("PCA on Health parameters")
print(health_pca)
summary(health_pca)
screeplot(health_pca, type="lines")
health_1<- health_pca$rotation[,1]
health_1
warning= FALSE
```

From the principle component analysis, the first two pc accounts for more than 80% of variations.

```{r health pc1 plot}
score_pc1_health<-health_1*health
sum_score_pc1_health<-rowSums(score_pc1_health[,1:3])
score_pc1_health["Sum_pc1"]<-sum_score_pc1_health
ggplot(score_pc1_health,aes(x=row.names(score_pc1_health),y=Sum_pc1))+ geom_col(mapping = NULL, data = NULL, position = "stack",
  width = NULL, na.rm = FALSE, show.legend = NA, inherit.aes = TRUE)+theme_bw()+geom_text(aes(label=row.names(score_pc1_health)),size=2,hjust=0.5,vjust=0)+ theme(legend.position="top",
              axis.text=element_text(angle=90, size = 6))
warning= FALSE
```
This plot is the scores of first pc of Health Matrix for each SSA country. SUM(pc1 * each variable)

```{r}
#subset equality parameters
G_inequal<-subset(data,Parameter.=="Gender inequality index")
G_inequal_melt<-melt(G_inequal,id.vars= c("Parameter.","Unit.","Source.","Year."))
G_inequal_melt$value<-as.numeric(G_inequal_melt$value)

#rural parameters
#safe drinkingwater
ac_dw_ru<-subset(data,Parameter.=="Rural population with access to safe drinking-water")
ac_dw_ru_melt<-melt(ac_dw_ru,id.vars=c("Parameter.","Unit.","Source.","Year."))
ac_dw_ru_melt$value<-as.numeric(ac_dw_ru_melt$value)
#clean cooking
ac_cc_ru<-subset(data,Parameter.=="Access to clean cooking in rural areas")
ac_cc_ru_melt<-melt(ac_cc_ru,id.vars=c("Parameter.","Unit.","Source.","Year."))
ac_cc_ru_melt$value<-as.numeric(ac_cc_ru_melt$value)

#urban parameters
#safe drinkingwater
ac_dw_ur<-subset(data,Parameter.=="Urban population with access to safe drinking-water")
ac_dw_ur_melt<-melt(ac_dw_ur,id.vars=c("Parameter.","Unit.","Source.","Year."))
ac_dw_ur_melt$value<-as.numeric(ac_dw_ur_melt$value)
#clean cooking
ac_cc_ur<-subset(data,Parameter.=="Access to clean cooking in urban areas")
ac_cc_ur_melt<-melt(ac_cc_ur,id.vars=c("Parameter.","Unit.","Source.","Year."))
ac_cc_ur_melt$value<-as.numeric(ac_cc_ur_melt$value)

#combine as a equality subset
equality<-cbind(G_inequal_melt$value, ac_dw_ru_melt$value, ac_cc_ru_melt$value, ac_dw_ur_melt$value, ac_cc_ur_melt$value)
equality<-as.data.frame(equality)
colnames(equality)<-c("G_inequal","ac_dw_rural","ac_cc_rural","ac_dw_urban","ac_cc_urban")
row.names(equality)<-Countries
equality_af<-head(equality,-5)
```




```{r}
#subset individual index and social index
data[grep("life",data$Parameter.,ignore.case=TRUE),]
lt_ppt<-subset(data,Parameter.=="Long-term annual precip depth")
tot_re_w<-subset(data,Parameter.=="Total renewable water sources")
dep_ra<-subset(data,Parameter.=="Dependency ratio")
var_interan<-subset(data,Parameter.=="Interannual variability")
var_seas<-subset(data,Parameter.=="Seasonal variability")
gdp<-subset(data,Parameter.=="Gross domestic product (GDP)")
gdp_pp<-subset(data,Parameter.=="GDP per capita")
edu_ind<-subset(data,Parameter.=="Education Index")
life_exp<-subset(data,Parameter.=="Life Expectancy at both sexs")
gdp<-gdp[1,]
hdi<-subset(data,Parameter.=="HDI")
tot_pop<-subset(data,Parameter.=="Population")

#melting dataframes for plotting purposes
lt_ppt_melt<-melt(lt_ppt,id.vars = c("Parameter.","Unit.","Source.","Year."))
lt_ppt_melt$value<-as.numeric(lt_ppt_melt$value)

tot_re_w_melt<-melt(tot_re_w,id.vars = c("Parameter.","Unit.","Source.","Year."))
tot_re_w_melt$value<-as.numeric(tot_re_w_melt$value)

dep_ra_melt<-melt(dep_ra,id.vars = c("Parameter.","Unit.","Source.","Year."))
dep_ra_melt$value<-as.numeric(dep_ra_melt$value)

var_interan_melt<-melt(var_interan,id.vars = c("Parameter.","Unit.","Source.","Year."))
var_interan_melt$value<-as.numeric(var_interan_melt$value)

var_seas_melt<-melt(var_seas,id.vars = c("Parameter.","Unit.","Source.","Year."))
var_seas_melt$value<-as.numeric(var_seas_melt$value)

gdp_melt<-melt(gdp,id.vars = c("Parameter.","Unit.","Source.","Year."))
gdp_melt$value<-as.numeric(gdp_melt$value)

gdp_pp_melt<-melt(gdp_pp,id.vars = c("Parameter.","Unit.","Source.","Year."))
gdp_pp_melt$value<-as.numeric(gdp_pp_melt$value)

edu_ind_melt<-melt(edu_ind,id.vars = c("Parameter.","Unit.","Source.","Year."))
edu_ind_melt$value<-as.numeric(edu_ind_melt$value)

life_exp_melt<-melt(life_exp,id.vars = c("Parameter.","Unit.","Source.","Year."))
life_exp_melt$value<-as.numeric(life_exp_melt$value)

hdi_melt<-melt(hdi,id.vars = c("Parameter.","Unit.","Source.","Year."))
hdi_melt$value<-as.numeric(hdi_melt$value)

tot_pop_melt<-melt(tot_pop,id.vars = c("Parameter.","Unit.","Source.","Year."))
tot_pop_melt$value<-as.numeric(tot_pop_melt$value)
#melt data needs to check if the row names are consistent do it tomorrow aug 2


#combine and construct dataframe, c means conbined value
lt_ppt_c<-lt_ppt_melt$value
tot_re_w_c<-tot_re_w_melt$value
dep_ra_c<-dep_ra_melt$value
var_interan_c<-var_interan_melt$value
var_seas_c<-var_seas_melt$value
gdp_c<-gdp_melt$value
hdi_c<-hdi_melt$value
gdp_pp_c<-gdp_pp_melt$value
edu_ind_c<-edu_ind_melt$value
life_exp_c<-life_exp_melt$value
ind_matrix<-cbind(lt_ppt_c, tot_re_w_c, dep_ra_c, var_interan_c, var_seas_c, gdp_c, hdi_c)
social_matrix<-cbind(gdp_pp_c, edu_ind_c, life_exp_c)
row.names(ind_matrix)<-Countries
row.names(social_matrix)<-Countries
ind_matrix_af<-head(ind_matrix,-5)
ind_matrix_5ref<-head(ind_matrix,5)
social_matrix_af<-head(social_matrix,-5)
social_matrix_5ref<-head(social_matrix,5)

#subset the access data
#data[grep("energy supply",data$Parameter.,ignore.case=TRUE),]
tot_pop_ac_dw<-subset(data, Parameter.=="Total population with access to safe drinking-water")
tot_pop_ac_dw_melt<-melt(tot_pop_ac_dw,id.vars = c("Parameter.","Unit.","Source.","Year."))
tot_pop_ac_dw_melt$value<-as.numeric(tot_pop_ac_dw_melt$value)
tot_pop_ac_dw_c<-tot_pop_ac_dw_melt$value

tot_pop_impr_sani<-subset(data, Parameter.=="Population with access to improved sanitation")
tot_pop_impr_sani_melt<-melt(tot_pop_impr_sani,id.vars = c("Parameter.","Unit.","Source.","Year."))
tot_pop_impr_sani_melt$value<-as.numeric(tot_pop_impr_sani_melt$value)
tot_pop_impr_sani_c<-tot_pop_impr_sani_melt$value
# Cut the redundancy of rural and urban ac_dw
# ac_matrix<-cbind(tot_pop_ac_dw_c, ac_dw_ur_melt$value, ac_dw_ru_melt$value, tot_pop_impr_sani_c,ener_pp_c, defi_food_c)
ac_matrix<-cbind(tot_pop_ac_dw_c, tot_pop_impr_sani_c, ener_pp_c, defi_food_c)

ac_matrix$value<-as.numeric(ac_matrix$value)
ac_matrix$`Depth of the food deficit`<-as.numeric(ac_matrix$`Depth of the food deficit`)
row.names(ac_matrix)<-Countries
#colnames(ac_matrix)<-c("tot_pop_ac_dw","ac_dw_ur", "ac_dw_ru", "tot_pop_impr_sani","ener_pp", "defi_food")
colnames(ac_matrix)<-c("tot_pop_ac_dw", "tot_pop_impr_sani","ener_pp", "defi_food")

ac_matrix_af<-head(ac_matrix,-5)
ac_matrix_5ref<-head(ac_matrix,5)
```

##PCA Analysis of Independent Variables
```{r}
#principle component analysis in independent variables
ind_matrix_pca<- ind_matrix_af[complete.cases(ind_matrix_af),]
ind_matrix_pca<- prcomp(ind_matrix_pca,center=TRUE, scale = TRUE)#matrix was normalized (scale=true) before perform pca analysis

print("PCA of Independent Variables")
print(ind_matrix_pca)
summary(ind_matrix_pca)
screeplot(ind_matrix_pca, type="lines")
ind_matrix_1<- ind_matrix_pca$rotation[,1]
ind_matrix_1
biplot(ind_matrix_pca)
```


The pca analysis for the independent variables  for 49 countries (data are available for most of the countries) was completed and the results was shown above. The first 4 principal components contributes 80 percent of the total variance. When looking at the first principal component, interannual variance and longtime precipitation are the top two variables who vary the most. In the second principal component, the seasonal variance and dependency ratio are the top two vairiables who vary the most. In the third principal component, the GDP is the predominent variable contributes to the total variance, while the total renewable water becomes the most important variable in the fourth principal component.


##PCA Analysis of Access Variables
```{r}
#principle component analysis in access variables
ac_matrix_pca<- ac_matrix_af[complete.cases(ac_matrix_af),]
ac_matrix_pca<- prcomp(ac_matrix_pca,center=TRUE, scale = TRUE)#matrix was normalized (scale=true) before perform pca analysis

print("PCA of access Variables")
print(ac_matrix_pca)
summary(ac_matrix_pca)
screeplot(ac_matrix_pca, type="lines")
ac_matrix_1<- ac_matrix_pca$rotation[,1]
ac_matrix_1
biplot(ac_matrix_pca)
```

The updated PC analysis for access matrix also have the kink. In the first pc, total population have access to safe drinking water and improved sanitation, and energy per capita were almost equally distributed (range from 0.49 to 0.56, positive sign); the deficit of food contributes less with a negative sign (-0.36). In the second pc, the predominent variable is deficit of food. Overall, pc1 and pc2 account for 73 percent of the total variation, pc3 will add another 16.4 percent.

```{r}
# Construct bar plots for each variable in Physical matrix
#long term precipitation
ind_matrix_af<-as.data.frame(ind_matrix_af)
#avg_lt_ppt_c<-mean(ind_matrix_af$lt_ppt_c)
ggplot(ind_matrix_af, aes(x=row.names(ind_matrix_af),y=lt_ppt_c))+ geom_col(mapping = NULL, data = NULL, position = "stack",
  width = NULL, na.rm = FALSE, show.legend = NA, inherit.aes = TRUE)+ theme(legend.position="top",
              axis.text=element_text(angle=90, size = 6))+ labs(x="Countries", y= "Long-term Precipitation (mm)")+ coord_cartesian(xlim= NULL,ylim = NULL, expand= TRUE)+geom_text(aes(label=row.names(ind_matrix_af)),size=2,hjust=0.5,vjust=0)
#+ geom_hline(yintercept = avg_lt_ppt_c)
warning= FALSE
hist(ind_matrix_af$lt_ppt_c, breaks = 20, xlab = "Long-Term precipitation (mm/year)", main = "Histogram: Long-Term Precipitation of SSA Countries")
```

```{r}
# Construct bar plots for each variable in Physical matrix
#total renewable water
#avg_lt_ppt_c<-mean(ind_matrix_af$lt_ppt_c)
ind_matrix_af$tot_re_w_c<-log10(ind_matrix_af$tot_re_w_c)
ggplot(ind_matrix_af, aes(x=row.names(ind_matrix_af), y=tot_re_w_c))+ geom_col(mapping = NULL, data = NULL, position = "stack",
  width = NULL, na.rm = FALSE, show.legend = NA, inherit.aes = TRUE)+ theme(legend.position="top",
              axis.text=element_text(angle=90, size = 6))+ labs(x="Countries", y= "Total Renewable Water Resources (10^9 cubic meter/year")+ coord_cartesian(xlim= NULL,ylim = NULL, expand= TRUE)+geom_text(aes(label=row.names(ind_matrix_af)),size=2,hjust=0.5,vjust=0)
#+ geom_hline(yintercept = avg_lt_ppt_c)
warning= FALSE
hist(ind_matrix_af$tot_re_w_c, breaks = 20, xlab="log10(total renewable water resources), log10(10^9 cubic meter per year)", main = "Histogram: Log10 of Total Renewable Water Resources of SSA Countries")
```

```{r}
# Construct bar plots for each variable in Physical matrix
#Dependency Ratio
#avg_lt_ppt_c<-mean(ind_matrix_af$lt_ppt_c)
ind_matrix_af$tot_re_w_c<-log10(ind_matrix_af$tot_re_w_c)
ggplot(ind_matrix_af, aes(x=row.names(ind_matrix_af), y=dep_ra_c))+ geom_col(mapping = NULL, data = NULL, position = "stack",
  width = NULL, na.rm = FALSE, show.legend = NA, inherit.aes = TRUE)+ theme(legend.position="top",
              axis.text=element_text(angle=90, size = 6))+ labs(x="Countries", y= "Dependency Ratio (%)")+ coord_cartesian(xlim= NULL,ylim = NULL, expand= TRUE)+geom_text(aes(label=row.names(ind_matrix_af)),size=2,hjust=0.5,vjust=0)
#+ geom_hline(yintercept = avg_lt_ppt_c)
warning= FALSE
hist(ind_matrix_af$dep_ra_c, breaks= 20, xlab = "Dependancy Ratio (%)", main = "Histogram: Dependancy ratio of SSA Countries")
```

Indicator expressing the percent of total renewable water resources originating outside the country. This indicator may theoretically vary between 0% and 100%. A country with a dependency ratio equal to 0% does not receive any water from neighbouring countries. A country with a dependency ratio equal to 100% receives all its renewable water from upstream countries, without producing any of its own. This indicator does not consider the possible allocation of water to downstream countries. 

```{r}
# Construct bar plots for each variable in Physical matrix
#interannual variability
#avg_lt_ppt_c<-mean(ind_matrix_af$lt_ppt_c)
ggplot(ind_matrix_af, aes(x=row.names(ind_matrix_af), y=var_interan_c))+ geom_col(mapping = NULL, data = NULL, position = "stack",
  width = NULL, na.rm = FALSE, show.legend = NA, inherit.aes = TRUE)+ theme(legend.position="top",
              axis.text=element_text(angle=90, size = 6))+ labs(x="Countries", y= "Interannual Variability")+ coord_cartesian(xlim= NULL,ylim = NULL, expand= TRUE)+geom_text(aes(label=row.names(ind_matrix_af)),size=2,hjust=0.5,vjust=0)
#+ geom_hline(yintercept = avg_lt_ppt_c)
warning= FALSE
hist(ind_matrix_af$var_interan_c, breaks = 20, xlab = "Interannual Variability (measured in standard deviations)", main = "Histogram: Interannual Viariability of SSA Countries")
```

This interannual variation was measured in the standard deviation and calculated using: "Standard deviation of annual total blue water divided by the mean of total blue water (1950–2008)".

```{r}
# Construct bar plots for each variable in Physical matrix
#seasonal variability
#avg_lt_ppt_c<-mean(ind_matrix_af$lt_ppt_c)
ggplot(ind_matrix_af, aes(x=row.names(ind_matrix_af), y=var_seas_c))+ geom_col(mapping = NULL, data = NULL, position = "stack",
  width = NULL, na.rm = FALSE, show.legend = NA, inherit.aes = TRUE)+ theme(legend.position="top",
              axis.text=element_text(angle=90, size = 6))+ labs(x="Countries", y= "Seasonal Variability")+ coord_cartesian(xlim= NULL,ylim = NULL, expand= TRUE)+geom_text(aes(label=row.names(ind_matrix_af)),size=2,hjust=0.5,vjust=0)
#+ geom_hline(yintercept = avg_lt_ppt_c)
warning= FALSE
hist(ind_matrix_af$var_seas_c, breaks = 20, xlab = "Seasonal Variability (measured in standard deviations)", main = "Histogram: Seasonal Viariability of SSA Countries")
```

This is the variability measured in standard deviation and calculated using: "standard deviation of monthly total blue water divided by the mean of monthly total blue water (1950–2008)."

```{r}
# Construct bar plots for each variable in Physical matrix
#GDP
#avg_lt_ppt_c<-mean(ind_matrix_af$lt_ppt_c)
log
ind_matrix_af$gdp_c<-log10(ind_matrix_af$gdp_c)
ggplot(ind_matrix_af, aes(x=row.names(ind_matrix_af), y=gdp_c))+ geom_col(mapping = NULL, data = NULL, position = "stack",
  width = NULL, na.rm = FALSE, show.legend = NA, inherit.aes = TRUE)+ theme(legend.position="top",
              axis.text=element_text(angle=90, size = 6))+ labs(x="Countries", y= "log10(GDP) (log10(million current USD)")+ coord_cartesian(xlim= NULL,ylim = NULL, expand= TRUE)+geom_text(aes(label=row.names(ind_matrix_af)),size=2,hjust=0.5,vjust=0)
#+ geom_hline(yintercept = avg_lt_ppt_c)
warning= FALSE
hist(ind_matrix_af$gdp_c, breaks = 20, xlab = "Log10(Gross Domestic Products) Log10(current million USD)", main = "Histogram: Log10 of GDP values of SSA Countries")
```


```{r}
#Construct bar plots for each variable in Physical matrix
#HDI
#avg_lt_ppt_c<-mean(ind_matrix_af$lt_ppt_c)
ggplot(ind_matrix_af, aes(x=row.names(ind_matrix_af), y=hdi_c))+ geom_col(mapping = NULL, data = NULL, position = "stack",
  width = NULL, na.rm = FALSE, show.legend = NA, inherit.aes = TRUE)+ theme(legend.position="top",
              axis.text=element_text(angle=90, size = 6))+ labs(x="Countries", y= "Human Development Index")+ coord_cartesian(xlim= NULL,ylim = NULL, expand= TRUE)+geom_text(aes(label=row.names(ind_matrix_af)),size=2,hjust=0.5,vjust=0)
#+ geom_hline(yintercept = avg_lt_ppt_c)
warning= FALSE
hist(ind_matrix_af$hdi_c)
```

```{r}
#plot HDI vs Seasonal Variation

ggplot(ind_matrix_af, aes(x=hdi_c, y=var_seas_c, color=row.names(ind_matrix_af)))+
  geom_point()+ geom_smooth(method="lm", color="black") +
  labs(x="Human Development Index", y= "Seasonal Variability")+
  theme(legend.position = "none")+ geom_text_repel(aes(hdi_c, var_seas_c, label = row.names(ind_matrix_af)))
warning= FALSE
```

As we see from the plot, HDI and Seasonal Variability is highly correlated. The decreasing trend can be found, which seasonal variability reduces as HDI increases. This is quite interesting and not really easy to explain. 

# Cluster Analysis for entire Independent Matrix

```{r}
#K means cluster analysis.
#remove na values and standarize the data
ind_matrix_af_ca<- na.omit(ind_matrix_af)
ind_matrix_af_ca<- scale(ind_matrix_af_ca)
#find out the number of rows where nigeria and south africa at.
#ind_matrix_af_ca<-cbind(ind_matrix_af_ca, as.numeric(1:41))

# Determine number of clusters
wss <- (nrow(ind_matrix_af_ca)-1)*sum(apply(ind_matrix_af_ca,2,var))
for (i in 2:15) wss[i] <- sum(kmeans(ind_matrix_af_ca, 
  	centers=i)$withinss)
plot(1:15, wss, type="b", xlab="Number of Clusters",
  ylab="Within groups sum of squares")

# K-Means Cluster Analysis
fit <- kmeans(ind_matrix_af_ca, 6) # 6 cluster solution
# get cluster means 
aggregate(ind_matrix_af_ca,by=list(fit$cluster),FUN=mean)
# append cluster assignment
ca_data <- data.frame(ind_matrix_af_ca, fit$cluster)

# Cluster Plot against 1st 2 principal components

# vary parameters for most readable graph

clusplot(ca_data, fit$cluster, color=TRUE, shade=TRUE, 
  	labels=2, lines=0, cex.txt=0.5)

# Centroid Plot against 1st 2 discriminant functions

plotcluster(ca_data, fit$cluster)
```

From the Cluster Analysis, in the plot of first two pc, we can see that the countries inside the same cluster are neighoring countries; some of the countries are in better economical standings. 

# Cluster Analysis for physical resources Parameters in the Independent Matrix. (Nigeria and South Africa were removed)

```{r}
#remove south africa and nigeria
num_countries<-cbind(row.names(ind_matrix_af_ca), as.numeric(1:nrow(ind_matrix_af_ca)))
colnames(num_countries)<-c("countries","order")
num_countries<-as.data.frame(num_countries)
num_countries$order<-as.numeric(1:nrow(num_countries))
nigeria<-subset(num_countries, countries =="Nigeria")
south_africa<-subset(num_countries, countries =="South.Africa")
nga<-nigeria$order
saf<-south_africa$order
ind_wo_N_SA<-ind_matrix_af_ca[-c(nga,saf),]
#K means cluster analysis.for the phsicial data
#remove na values and standarize the data

#subset physical data
ind_wo_N_SA_ca_ph<-ind_wo_N_SA[,-c(6,7)]

# Determine number of clusters
wss <- (nrow(ind_wo_N_SA_ca_ph)-1)*sum(apply(ind_wo_N_SA_ca_ph,2,var))
for (i in 2:15) wss[i] <- sum(kmeans(ind_wo_N_SA_ca_ph, 
  	centers=i)$withinss)
plot(1:15, wss, type="b", xlab="Number of Clusters",
  ylab="Within groups sum of squares")

# K-Means Cluster Analysis
fit <- kmeans(ind_wo_N_SA_ca_ph, 5) # 5 cluster solution
# get cluster means 
aggregate(ind_wo_N_SA_ca_ph,by=list(fit$cluster),FUN=mean)
# append cluster assignment
ca_data <- data.frame(ind_wo_N_SA_ca_ph, fit$cluster)

# Cluster Plot against 1st 2 principal components

# vary parameters for most readable graph
 
clusplot(ca_data, fit$cluster, color=TRUE, shade=TRUE, 
  	labels=2, lines=0, cex.txt=0.5)

# Centroid Plot against 1st 2 discriminant functions

plotcluster(ca_data, fit$cluster)
```

# Cluster Analysis for Social and Economical Parameters in the Independent Matrix.

```{r}
#subset social-economic data
ind_wo_N_SA_ca_se<-ind_wo_N_SA[,c(6,7)]
#Need to address the discrepency issues btw rmd and knit files
# Determine number of clusters
wss <- (nrow(ind_wo_N_SA_ca_se)-1)*sum(apply(ind_wo_N_SA_ca_se,2,var))
for (i in 2:15) wss[i] <- sum(kmeans(ind_wo_N_SA_ca_se, 
  	centers=i)$withinss)
plot(1:15, wss, type="b", xlab="Number of Clusters",
  ylab="Within groups sum of squares")

# K-Means Cluster Analysis
fit_se <- kmeans(ind_wo_N_SA_ca_se, 6) # 6 cluster solution
# get cluster means 
aggregate(ind_wo_N_SA_ca_se,by=list(fit_se$cluster),FUN=mean)
# append cluster assignment
ca_data_se<- data.frame(ind_wo_N_SA_ca_se, fit_se$cluster)

# Cluster Plot against 1st 2 principal components

# vary parameters for most readable graph

clusplot(ca_data_se, fit_se$cluster, color=TRUE, shade=TRUE, 
  	labels=2, lines=0, cex.txt=0.5)

# Centroid Plot against 1st 2 discriminant functions

plotcluster(ca_data_se, fit_se$cluster)
```

```{r}
#remove south africa and nigeria
num_countries_sm<-cbind(row.names(social_matrix_af), as.numeric(1:nrow(social_matrix_af)))
colnames(num_countries_sm)<-c("countries","order")
num_countries_sm<-as.data.frame(num_countries_sm)
num_countries_sm$order<-as.numeric(1:nrow(num_countries_sm))
nigeria_sm<-subset(num_countries_sm, countries =="Nigeria")
south_africa_sm<-subset(num_countries_sm, countries =="South.Africa")
nga_sm<-nigeria_sm$order
saf_sm<-south_africa_sm$order
social_matrix_af<-social_matrix_af[-c(nga_sm,saf_sm),]
#principle component analysis in independent variables
social_matrix_pca<- social_matrix_af[complete.cases(social_matrix_af),]
social_matrix_pca<- prcomp(social_matrix_pca,center=TRUE, scale = TRUE)#matrix was normalized (scale=true) before perform pca analysis

print("PCA of social Variables")
print(social_matrix_pca)
summary(social_matrix_pca)
screeplot(social_matrix_pca, type="lines")
social_matrix_1<- social_matrix_pca$rotation[,1]
social_matrix_1
biplot(social_matrix_pca)
```

```{r}
#K means cluster analysis.
#remove na values and standarize the data
social_matrix_af_ca<- na.omit(social_matrix_af)
social_matrix_af_ca<- scale(social_matrix_af_ca)
#find out the number of rows where nigeria and south africa at.
#ind_matrix_af_ca<-cbind(ind_matrix_af_ca, as.numeric(1:41))

# Determine number of clusters
wss <- (nrow(social_matrix_af_ca)-1)*sum(apply(social_matrix_af_ca,2,var))
for (i in 2:15) wss[i] <- sum(kmeans(social_matrix_af_ca, 
  	centers=i)$withinss)
plot(1:15, wss, type="b", xlab="Number of Clusters",
  ylab="Within groups sum of squares")

# K-Means Cluster Analysis
fit <- kmeans(social_matrix_af_ca, 6) # 6 cluster solution
# get cluster means 
aggregate(social_matrix_af_ca,by=list(fit$cluster),FUN=mean)
# append cluster assignment
ca_data <- data.frame(social_matrix_af_ca, fit$cluster)

# Cluster Plot against 1st 2 principal components

# vary parameters for most readable graph

clusplot(ca_data, fit$cluster, color=TRUE, shade=TRUE, 
  	labels=2, lines=0, cex.txt=0.5)

# Centroid Plot against 1st 2 discriminant functions

plotcluster(ca_data, fit$cluster)
```
